---
title: "final_project"
author: "Tiana Bell"
date: "8/10/2018"
output: html_document
---

```{r setup, include=FALSE}
## install packages if necessary
list.of.packages <- c("janitor", "ggplot2", "dplyr", "lubridate", "zoo", "forcats", "ggthemes", "knitr", "rmarkdown", "nanair")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## load packages
library(janitor)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)
library(forcats)
library(ggthemes)
library(knitr)
library(naniar)

## set working directory
knitr::opts_knit$set(root.dir = '/cloud/project/final_project')
```

### The Data

The American Time Use Survey (ATUS) is a time-use survey of Americans, which is sponsored by the Bureau of Labor Statistics (BLS) and conducted by the U.S. Census Bureau. Respondents of the survey are asked to keep a diary for one day carefully recording the amount of time they spend on various activities including working, leisure, childcare, and household activities. The survey has been conducted every year since 2003.

Included in the data are main demographic variables such as respondents' age, sex, race, marital status, and education. The data also includes detailed income and employment information for each respondent. While there are some slight changes to the survey each year, the main questions asked stay the same. You can find the data dictionaries for each year on [https://www.bls.gov/tus/dictionaries.htm](https://www.bls.gov/tus/dictionaries.htm)


### Accessing the Data

There are multiple ways to access the ATUS data; however, for this project, you'll get the raw data directly from the source. The data for each year can be found at [https://www.bls.gov/tus/#data](https://www.bls.gov/tus/#data). Once there, there is an option of downloading a multi-year file, which includes data for all of the years the survey has been conducted, but **for the purposes of this project, let's just look at the data for 2016**. Under **Data Files**, click on `American Time Use Survey--2016 Microdata files`. 

You will be brought to a new screen. Scroll down to the section **2016 Basic ATUS Data Files**. Under this section, you'll want to **click to download** the following two files: `ATUS 2016 Activity summary file (zip)` and `ATUS-CPS 2016 file (zip)`. 

* `ATUS 2016 Activity summary file (zip)` contains information about the total time each ATUS respondent spent doing each activity listed in the survey. The activity data includes information such as activity codes, activity start and stop times, and locations.
* `ATUS-CPS 2016 file (zip)` contains information about each household member of all individuals selected to participate in the ATUS.

Once they've been downloaded, you'll need to **unzip the files**. Once unzipped, you will see the dataset in a number of different file formats including `.sas`, `.sps`, and `.dat` files. **We'll be working with the .dat files.**

### Loading the Data into R

Use the first approach explained above to download and access the ATUS data for 2016. Download the CPS and Activity Summary files in a folder and unzip them and within each folder upload the files ending in .dat to `data/raw_data` filder on RStudio.cloud. To load the data in, **run the code in the `atus-data` code chunk** to create an object called `atus.all`.

### Importing data

```{r atus-data, warning=FALSE}
atus.cps <- read.delim('/cloud/project/final_project/data/raw_data/atuscps_2016.dat', sep=",")
atus.sum <- read.delim('/cloud/project/final_project/data/raw_data/atussum_2016.dat', sep=",")
atus.all <- atus.sum %>%
  left_join(atus.cps %>% filter(TULINENO==1), by = c("TUCASEID"))
```

### Exploratory Analysis of Child Care Data

```{r initial-exploration}
### Add Code Here
str(atus.all) 
summary(atus.all)
colnames(atus.all)
rownames(atus.all)
 ## find average amount of time person spends doing "t120101"
mean(atus.all$t120101)

```


```{r creating-childcare-var}

## write out al columns up to t030112 for sum of CHILDCARE column.
atus.all <- atus.all %>% 
  mutate(CHILDCARE = t030101 + t030102 + t030103 + t030104 + t030105 + t030106 + t030108 + t030109 + t030110 + t030111 + t030112 %>%
           glimpse(CHILDCARE))
```

```{r childcare-density-plot}
ggplot(atus.all, aes(CHILDCARE, na.rm = FALSE)) +
    geom_density()
```
This graph shows us the respondents that correlate with the amount of time spent with their children.
```{r gender-analysis}
atus.all %>% 
    group_by(TESEX) %>% 
   summarise(avg_parent_childcare = mean(CHILDCARE))
```
In this analysis we can tell that males(variable = 1) spend on average 19 mins with children compared to females (variable = 2) that spend on average 33 minutes with children. These numbers are a reflection of individual respondents without other factors(i.e. marital status, age, etc.).
```{r replacing-na}
## replace -1 in the variable TRDPFTPT with NA.
  atus.all$TRDPFTPT[atus.all$TRDPFTPT == -1] <- NA %>%
  sum(is.na(atus.all$TRDPFTPT))##take this away, filter once done. 
grep("TRHHCHILD",names(atus.all))
#find the amount of missing values in the column
sum(is.na(atus.all$TRDPFTPT))
class(atus.all$TRYHHCHILD)
```


```{r exploratory-analysis}
## add your exploratory analysis code here
## do younger people spend more time with children than older people.
adults_atLeast_one_kid <- atus.all %>%
  select(CHILDCARE, TEAGE, TRYHHCHILD, HEFAMINC, TRCHILDNUM, PEMARITL, TRDPFTPT) %>%
       filter(TRCHILDNUM >  0)
 ggplot(adults_atLeast_one_kid, aes(x = TEAGE, y = CHILDCARE)) +
      geom_point()
```
 
```{r exploratory-analysis 2}
## do richer people spend more time with children than poor
      ggplot(adults_atLeast_one_kid, aes(x = HEFAMINC, y = CHILDCARE)) +
      geom_point() +
      labs( x = "FAMILY INCOME \n tens of thousands", y = "CHILDCARE \n minutes per week", title = "Do richer people spend more time with \n their children than poorer people?")
```


```{r exploratory-analysis 3}
## do married people spend more time with children than single parents
      ggplot(adults_atLeast_one_kid, aes(x = PEMARITL, y = CHILDCARE)) +
        geom_point()
```

``` {r exploratory-analysis 4}
## do full-time workers spend more time than part-time workers?
      ggplot(adults_atLeast_one_kid, aes(x = TRDPFTPT, y = CHILDCARE)) +
        geom_point()

##make ggplot y axis is amount of time spent, x axis the age
##y axis is on the left hand side, x axis is on the right hand side; time spent variable would be on the left hand side
```

### Regression Analysis

```{r reg-analysis}
## add your regression analysis code here
young_vs_old <- lm(CHILDCARE ~ TEAGE + HEFAMINC + PEMARITL +, data = adults_atLeast_one_kid)
summary(young_vs_old)
```

### Exploratory Analysis of Age and Activities

```{r activity-cats}
## must select column that reflects the row I would like to work in
## select 
atus.wide <- atus.all %>%
    mutate(act01 = rowSums(atus.all[,grep("t01", names(atus.all))]),
           act02 = ,
           act03 = ,
           act04 = ,
           act05 = ,
           act06 = ,
           act07 = ,
           act08 = ,
           act09 = ,
           act10 = ,
           act11 = ,
           act12 = ,
           act13 = ,
           act14 = ,
           act15 = ,
           act16 = ,
           # act17 = , there is no category 17 in the data
           act18 = ) %>% 
    select(TUCASEID, TEAGE, HEFAMINC, starts_with("act"))
```

```{r wide-to-long}
atus.long <- atus.wide %>% 
  # use code to convert the wide format to long.
```

```{r age-activity}
atus.long %>% 
    group_by(ACTIVITY, TEAGE) %>% 
    summarise(AVGMINS = mean(MINS)) %>% 
    ggplot() +
```

### Exploratory Analysis of Income and Activities

```{r activity-income}
atus.long %>% 
  group_by(ACTIVITY, HEFAMINC) %>% 
  ## add the rest of the code here
```


```{r save-plot}
## save the plot above
```